---
title: 'ECON312 Problem Set 5'
author: Futing Chen, Hongfan Chen, Will Parker
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_notebook:
    toc: yes
    toc_depth: 2
    toc_float: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, cache = FALSE}
library(dplyr)
library(knitr)
library(haven)
```


# 1. Describe the data
```{r}
data <- read_dta("PS5.dta")
psych::describe(data) %>% 
  select(n, mean, sd, median, min, max) %>%
  kable(digits = 2)
```

# 2. Estimate the following regression on the sample of fast food restaurants in Feb-Mar 1992:

$$empft_{ikt} = \alpha + \gamma minwage_{kt} + \beta_1 nregs_{ikt} + \beta_2 hrsopen_{ikt} + \sum_{j=2}^4 \eta_j d_j + \epsilon_{ikt}$$
$i$ denotes restaurant, $k$ denotes state, and $t = 0$ if the observation is from Feb-Mar and $t = 1$ if the observation is from Nov-Dec.

```{r, results='asis'}
# drop missing observations
for_regression <- data %>%
  select(empft, minwage,nregs, hrsopen, d2, d3, d4) %>%
  na.omit()
regress_formula <- formula(empft ~ minwage + nregs + hrsopen + d2 + d3 + d4)
linear_model <- lm(regress_formula, data = for_regression)
stargazer::stargazer(linear_model, header = FALSE)
```

# 3. Interpret the coefficient $\gamma$ and calculate a 90% confidence interval.

$\gamma =$ `r round(linear_model$coefficients['minwage'], 2) %>% as.numeric()` is the average change in number of full time employees for a \$1 increase in the minimum wage, adjusted for the other variables in the regression. The 90% CI is
```{r}
confint.lm(linear_model, level = 0.9)[2,]
```

# 4.Use the Sum of squares table from the regression output to calculate the $R^2$ and the standard error of the regression (Root MSE).

```{r}
SSE <- sum(linear_model$residuals^2)
SST <- var(for_regression$empft)*(for_regression %>% nrow() -1)
SSR <- SST - SSE
R2 <- SSR/SST
root_MSE <- sqrt(SSE/(for_regression %>% nrow()))
```

The $R^2 =$ `r round(R2, 2)` and the $MSE =$ `r round(root_MSE, 1)`

# 5. Give an economic interpretation of the coefficients $\eta_2, \eta_3, \eta_4$. What might explain the relatively large coefficient on -d4-?

These are fixed effects for each restaurant, specifically the average difference in number of employees relative to burger king. d-1 The large coefficient $\eta_4$ means that Wendy's employed `r round(linear_model$coefficients['d4'], 2) %>% as.numeric()` more people on average than Burger King, adjusted for the other variables in the regression. One possible explanation is that Wendy's has more customers and hence hires more employers.

# 6. Test $H0: \eta_2 = \eta_3 = 0$
```{r}
car::linearHypothesis(linear_model, c("d2 = 0", "d3 = 0"))
```
We can reject the hypothesis $\eta_2 = \eta_3 = 0$ at a significane level of $\alpha = 0.5$

# 7. Test the hypothesis $H_0 : \eta_2 = \eta_3$ using the estimated covariance matrix of the coefficients. Verify your answer by running the test in Stata using and/or by performing an F-test. We now want to control for potential selection issues by using the panel structure of our data.

```{r}
cov_matrix <- vcov(linear_model)
beta <- linear_model$coefficients
R <- c(0,0,0, 0, 1, -1, 0)
n_sample <- for_regression %>% nrow()
t(R)%*%beta
```

```{r}
T_n <- n_sample*(R%*%beta)%*%solve(t(R)%*%cov_matrix%*%R)%*%t(R%*%beta)
```

```{r}
car::linearHypothesis(linear_model, c("d2 = d3"))
```

# 8. Explain why the previous estimate of $\lambda$ is likely to suffer from omitted variable bias.
$\lambda$ suffer from omitted variable bias if some of the unobservables that affect the number of employees are correlated with minimum wage. The omitted variable bias is likely because one can imagine many such unobservables such as unmeasured conditions of state economies or demographics. For example, some states have higher per capita income or the age structure is younger. These characteristics are potentially positively correlated with $minwage$ and $empft$; then $\gamma$ is upwardly biased.

# 9. Assume that $\epsilon_{ikt} = \mu_k + \zeta_t + u_{ikt}$ and that $E[u_{ikt}|X_{ikt}] = 0$ (where X_{ikt} is the vector of RHS-variables in (2) except -minwage-). Explain how you can then use the increase in the minimum wage in New Jersey and a difference-in-differences (DD) model to identify the effect of the minimum wage on employment. Give an example where the necessary assumption(s) are violated.
Since $\epsilon_{ikt} = \mu_k + \zeta_t + u_{ikt}$, we know
\[Y=\alpha+\gamma minwage_{kt}+X_{ikt}^{'}\beta+\mu_k+\zeta_t+u_{ikt} \]
Then $E[u_{ikt}|X_{ikt}] = 0$ implies
\begin{align*}
E[Y|post=1, state=1, X_{ikt}]-E[Y|post=0, state=1, X_{ikt}]&=\gamma \Delta minwage+(\zeta_1-\zeta_0)\\
E[Y|post=1, state=0, X_{ikt}]-E[Y|post=0, state=0, X_{ikt}]&=(\zeta_1-\zeta_0)
\end{align*}
Thus, 
\begin{align*}
&\{E[Y|post=1, state=1, X_{ikt}]-E[Y|post=0, state=1, X_{ikt}]\} \\
-&\{E[Y|post=1, state=0, X_{ikt}]-E[Y|post=0, state=0, X_{ikt}]\}=\gamma \Delta minwage
\end{align*}
Hence, we can use the following DD model:
\[empft_{ikt} =\pi_0+\delta_{kt} state_{k}*post_t+ \pi_1state_{k}+\pi_2post_t+X_{ikt}^{'}\beta + u_{ikt}\]
and the DD estimator $\delta_{kt}$ will identify the effect of the minimum wage on employment.\par
The necessary assumption is common trend in the absence of intervation:
\[E[Y_1^0-Y_0^0|D_1=1]=E[Y_1^0-Y_0^0|D_1=0]\]
This would be  violated if, for example, the employment of one state (say, New Jersey) benefited more from progressive federal policies than Pennsylvania.

# 10. Generate a table of means, a table of standard errors and a table of frequencies for -empft- in each state and each time period (post = 1 and post = 0).
```{r}
with(data, tapply(empft, list(state=state,post=post), mean,na.rm=T) )
with(data, tapply(empft, list(state=state,post=post), sd,na.rm=T) )
filter(data, !is.na(empft)) %>% 
  with(., tapply(empft, list(state=state,post=post), length) )
```

# 11. Using these statistics, calculate a DD estimate of the impact of the minimum wage law on employment.

```{r}
(8.446875-7.732308)-(7.651316-10.311688)
```

# 12. Specify and estimate the corresponding regression.
The regression corresponding to (11) is 
\[empft_{ikt}=\pi_0+\delta state_{k}*post_t+\pi_1state_{k}+\pi_2post_t+v_{ikt}\]
```{r}
DID1<-lm(empft~state*post,data=data); summary(DID1)
```

We can also add covariates and estimate
```{r}
DID2<-lm(empft~state*post+nregs+hrsopen+d2+d3+d4,data=data); summary(DID2)
```

# 13. How much does this suggest that the minimum wage affects full time employment in fast food restaurants?
```{r}
DID_est<-coef(DID1)[4]
names(DID_est)<-NULL
DID_est/(5.05-4.25)
```
The DID estimate is $3.3749$. Dividing this by $\Delta minwage$, we obtain $4.219$, meaning that increasing minimum wage by 1 dollar will increase the number of employees in fast food restaurants by 4.
