---
title: 'ECON312 Problem Set 5'
author: Futing Chen, Hongfan Chen, Will Parker
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_notebook:
    toc: yes
    toc_depth: 2
    toc_float: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, cache = FALSE}
library(tidyverse)
library(knitr)
library(haven)
```


# 1. Describe the data
```{r}
data <- read_dta("PS5.dta")

psych::describe(data) %>% 
  select(n, mean, sd, median, min, max) %>%
  kable(digits = 2)
```

# 2. Estimate the following regression on the sample of fast food restaurants in Feb-Mar 1992:

$$empft_{ikt} = \alpha + \gamma minwage_{kt} + \beta_1 nregs_{ikt} + \beta_2 hrsopen_{ikt} + \sum_{j=2}^4 \eta_j d_j + \epsilon_{ikt}$$
$i$ denotes restaurant, $k$ denotes state, and $t = 0$ if the observation is from Feb-Mar and $t = 1$ if the observation is from Nov-Dec.

```{r, results='asis'}
# drop missing observations
for_regression <- data %>%
  select(empft, minwage,nregs, hrsopen, d2, d3, d4) %>%
  na.omit()

regress_formula <- formula(empft ~ minwage + nregs + hrsopen + d2 + d3 + d4)

linear_model <- lm(regress_formula, data = for_regression)

stargazer::stargazer(linear_model, header = FALSE)
```

# 3. Interpret the coefficient $\gamma$ and calculate a 90% confidence interval.

$\gamma =$ `r round(linear_model$coefficients['minwage'], 2) %>% as.numeric()` is the average change in number of full time employees for a \$1 increase in the minimum wage, adjusted for the other variables in the regression. The 90% CI is
```{r}
confint.lm(linear_model, level = 0.9)[2,]
```

# 4.Use the Sum of squares table from the regression output to calculate the $R^2$ and the standard error of the regression (Root MSE).

```{r}
SSE <- sum(linear_model$residuals^2)
SST <- var(for_regression$empft)*(for_regression %>% nrow() -1)
SSR <- SST - SSE

R2 <- SSR/SST
root_MSE <- sqrt(SSE/(for_regression %>% nrow()))
```

The $R^2 =$ `r round(R2, 2)` and the $MSE =$ `r round(root_MSE, 1)`

# 5. Give an economic interpretation of the coefficients $\eta_2, \eta_3, \eta_4$. What might explain the relatively large coefficient on -d4-?

These are fixed effects for each restaurant, specifically the average difference in number of employees relative to burger king. d-1 The large coefficient $\eta_4$ means that Wendy's employed `r round(linear_model$coefficients['d4'], 2) %>% as.numeric()` more people on average than Burger King, adjusted for the other variables in the regression.

# 6. Test $H0: \eta_2 = \eta_3 = 0$
```{r}
car::linearHypothesis(linear_model, c("d2 = 0", "d3 = 0"))
```
We can reject the hypothesis $\eta_2 = \eta_3 = 0$ at a significane level of $\alpha = 0.5$

# 7. Test the hypothesis $H_0 : \eta_2 = \eta_3$ using the estimated covariance matrix of the coefficients. Verify your answer by running the test in Stata using and/or by performing an F-test. We now want to control for potential selection issues by using the panel structure of our data.

```{r}


cov_matrix <- vcov(linear_model)

beta <- linear_model$coefficients

R <- c(0,0,0, 0, 1, -1, 0)

n_sample <- for_regression %>% nrow()

t(R)%*%beta
```

```{r}
T_n <- n_sample*(R%*%beta)%*%solve(t(R)%*%cov_matrix%*%R)%*%t(R%*%beta)
```

```{r}
car::linearHypothesis(linear_model, c("d2 = d3"))
```

# 8. Explain why the previous estimate of $\lambda$ is likely to suffer from omitted variable bias.

