---
title: 'ECON312 Problem Set B3'
author: Futing Chen, Hongfan Chen, Will Parker
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_notebook:
    toc: yes
    toc_depth: 2
    toc_float: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, cache = FALSE}
library(tidyverse)
library(knitr)
library(latex2exp)

set.seed(112345)
```


# 2: Roy Model of College

$$Y_1 =  \alpha + \bar{\beta} + U_1$$
$$Y_0 = \alpha + U_0$$

$$D = \mathbf{1}\{Y_1 - Y_0 - C \geq 0\}$$

$$\begin{pmatrix} U_0 \\ U_1 \\ V \end{pmatrix} \sim N(0, \begin{bmatrix} \sigma^2_0 & \sigma_{01} & \sigma_{0,V} \\ \sigma_{01} & \sigma_1^2 & \sigma_{1, V} \\ \sigma_{0,V} & \sigma_{1, V} & \sigma_V^2 \end{bmatrix}) $$

Where $C = Z + V$, $Z$ is tuition and $V$ is psychic costs

$$Z \perp \!\!\! \perp (U_1, U_0, V)$$

Where $\alpha = 0.67$ and $\bar{\beta} = 0.2$

Assume $Z \sim U[-0.5, 0.5]$

## Monte Carlo simulation of model
```{r}

roy_MC <- function(config_num, sigma_11, sigma_00, sigma_V, sigma_01, sigma_0V, sigma_1V,
                   alpha = 0.67, beta = 0.2, N = 100000,
                   zmin = -0.5, zmax = 0.5
                   ){
  
  Z <- runif(N, zmin, zmax)
  
  if (sigma_V > 0){
      Sigma <- matrix(c(sigma_00, sigma_01, sigma_0V, sigma_01, sigma_11, sigma_1V, sigma_0V, sigma_1V, sigma_V^2), nrow = 3)
      
        latent_vars <- MASS::mvrnorm(N, c(0,0,0), Sigma) %>%
           as_tibble()
        
        colnames(latent_vars) <- c("U_0", "U_1", "V")
   
  } else {
          Sigma <- matrix(c(sigma_00, sigma_01, sigma_01, sigma_11), nrow = 2)
          
        latent_vars <- MASS::mvrnorm(N, c(0,0), Sigma) %>%
           as_tibble()
        
        colnames(latent_vars) <- c("U_0", "U_1")
        
        latent_vars <- latent_vars %>%
          mutate(V = 0)
          
  }

  
  sigma_v_tilde <- sqrt(sigma_00 + sigma_11 + sigma_V -2*sigma_01 +2*sigma_0V - 2*sigma_1V)
  
  
  outcomes <- latent_vars %>%
    cbind(Z) %>%
    mutate(Y_1 = alpha + beta + U_1,
           Y_0 = alpha + U_0,
           beta_i = Y_1 - Y_0,
           C = Z + V,
           D = ifelse(Y_1 - Y_0 - C >= 0, 1, 0),
           V_tilde = V + U_0 - U_1,
           U_D = pnorm(V_tilde/sigma_v_tilde),
           MTE = beta + V_tilde*(-sigma_11 - sigma_00 +2*sigma_01 +sigma_1V - sigma_0V)/sigma_v_tilde^2,
           Y = D*Y_1 + (1-D)*Y_0,
           config_num = config_num)
  
  outcomes
  
}


```


### Simulations for $Z ~ U(-0.5, 0.5)$
```{r}
config_1 <- roy_MC("I", 1,1,1,0,0,0)
config_2 <- roy_MC("II",1,1,1,0.5,-0.5,0)
config_3 <- roy_MC("III",1,1,1,0,0,0.5)
config_4 <- roy_MC("IV",1,1,0, 0.5,0,0)
config_5 <- roy_MC("V", 1,0.25,1,0.1,-0.2,0.2)
config_6 <- roy_MC("VI",0.25,0.25,1,-0.2,0.1,-0.2)
config_7 <- roy_MC("VII",1,1,1,-0.2,0.6,0.5)


config_list <- list(config_1, config_2, config_3, 
                    config_4, config_5, config_6, config_7)
```


## A. Derive and graph the MTE

The propensity score $P(Z)$ is

$$P(D = 1| Z) = E[\mathbf{1}\{Y_1 - Y_0 - C \geq 0\}|Z]$$

plugging in for $Y_1 - Y_0 - C = \bar{\beta} + U_1 - U_0 - Z - V$, we have


$$P(Z) = P(\bar{\beta} + U_1 - U_0 - Z - V \geq 0)$$
re-arranging we have

$$P(Z) = P(\bar{\beta}  - Z \geq V + U_0 - U_1)$$
Or

$$P(Z) = P( V + U_0 - U_1 \leq \bar{\beta}  - Z)$$

Writing it this way, we can define a latent variable $\tilde{V} = -U_1 + U_0 + V$ that will contribute to treatment choice. The Computing the variance $Var(-U_1 + U_0 + V)$, we have


$$Var(-U_1 + U_0 + V) = \sigma_{\tilde{V}}^2 =  \sigma_v^2 + \sigma_0^2 + \sigma_1^2 - 2\sigma_{01} + 2\sigma_{0V} -2\sigma_{1V}$$

Since we know that $(U_1, U_0, V)$ is distributed as multivariate normal, $\tilde{V} \sim N(0, \sqrt{\sigma_v^2 + \sigma_0^2 + \sigma_1^2 - 2\sigma_{01} + 2\sigma_{0V} -2\sigma_{1V}})$. So we can re-write the propensity score as

$$P(D=1|Z) = P(Z =z) = P(\frac{\tilde{V}}{\sigma_{\tilde{V}}} \leq \frac{\bar{\beta}  - Z}{\sigma_{\tilde{V}}}) $$
and since $\frac{\tilde{V}}{\sigma_{\tilde{V}}} \sim N(0,1)$, finally we can write 

$$P(Z) = U_D = \Phi(\frac{\bar{\beta}  - Z}{\sigma_{\tilde{V}}}) $$

Therefore we define the MTE as

$$MTE(U_D) = E[Y_1 - Y_0 | U_D = u_D] = E[Y_1 - Y_0 | \tilde{V} = \tilde{v}] $$
Because of **index sufficiency**. 

$$E[Y_1 -Y_0| \tilde{V} = \tilde{v}] = \bar{\beta} + E[U_1 -U_0| \tilde{V} = \tilde{v}]$$
Because of the given multivariate normal relationship, we can expand the second term

$$E[Y_1 -Y_0| Z = z ] = \bar{\beta} + Cov(U_1 - U_0, \frac{-U_1 + U_0 + V}{\sigma_{\tilde{V}}})*(-U_1 + U_0 + V)$$

Note

$$Cov(U_1 - U_0,\frac{-U_1 + U_0 + V}{\sigma_{\tilde{V}}}) = \frac{1}{\sigma_{\tilde{V}}}(-Var(U_1 - U_0) + Cov(U_1 - U_0, V)) = \frac{1}{\sigma_{\tilde{V}}}(-\sigma_1^2 - \sigma_0^2 + 2\sigma_{01} + \sigma_{V1} - \sigma_{V0}) $$
So plugging in we have 


$$\Delta^{MTE} = \bar{\beta} + \frac{-\sigma_1^2 - \sigma_0^2 + 2\sigma_{01} + \sigma_{V1} - \sigma_{V0}}{ \sigma_v^2 + \sigma_0^2 + \sigma_1^2 - 2\sigma_{01} + 2\sigma_{0V} -2\sigma_{1V}}*(V + U_0 -U_1)$$



```{r}

MTE <- function(config_num,
                     sigma_11, sigma_00, sigma_V, sigma_01, sigma_0V, sigma_1V,
                     beta = 0.2){
  
  sigma_v_tilde <- sqrt(sigma_00 + sigma_11 + sigma_V -2*sigma_01 +2*sigma_0V - 2*sigma_1V)
  
  V_tilde <- seq(-6, 6, 0.1)
  
  tibble(V_tilde) %>%
    mutate(configuration = config_num,
           U_d = pnorm(V_tilde/sigma_v_tilde),
           MTE = beta + V_tilde*(-sigma_11 - sigma_00 +2*sigma_01 +sigma_1V - sigma_0V)/sigma_v_tilde^2)
  
}

all_MTEs <- 
  MTE("I", 1,1,1,0,0,0) %>%
  rbind(MTE("II", 1,1,1,0.5,-0.5,0),
        MTE("III", 1,1,1,0,0,0.5),
        MTE("IV", 1,1,0, 0.5, 0, 0),
        MTE("V", 1, 0.25, 1, 0.1, -0.2, 0.2),
        MTE("VI", 0.25, 0.25, 1, -0.2, 0.1, -0.2),
        MTE("VII", 1, 1, 1, -0.2, 0.6, 0.5)) 
```


```{r}
all_MTEs %>%
  ggplot(aes(x = V_tilde, y = MTE, color = configuration)) +
  geom_line() + labs(x = TeX("V + U_0 - U_1"), y = TeX("$\\Delta^{MTE}$"), linetype = TeX("$\\bar{\\beta}$")) + geom_hline(aes(linetype = " ", yintercept = 0.2))
```

Alternatively with the latent variable normalized


## Compute and plot LIV estimates for the configurations of parameters and data from different societies listed below.

The Local IV is defined to be

$$\Delta^{LIV}(P(Z)) = \frac{\partial E[Y|P(Z) = P(z)]}{\partial P(z)} = E[Y_1 - Y_o|P(Z) = P(z)]$$

So we ccan write the LIV as a function of the MTE

$$\Delta^{LIV}(P(Z)) = MTE(U_D = P(z)) $$


```{r}
all_MTEs %>%
  ggplot(aes(x = U_d, y = MTE, color = configuration)) +
  geom_line() + labs(x = TeX("$U_D = P(Z)$"), y = TeX("$\\Delta^{LIV}$"), linetype = TeX("$\\bar{\\beta}$")) + geom_hline(aes(linetype = " ", yintercept = 0.2))
```

### Verifying result with MC simulations
```{r}
config_1 %>%
  ggplot(aes(x = V_tilde))+ 
  geom_smooth(aes(y = beta_i, color = "MTE (from MC simulation)"), method = "lm") +
  geom_point(aes(y = beta_i), alpha = 0.1, size = 0.1) +
  geom_line(aes(y = MTE, color = "MTE (theory)")) + labs(color = "") +
  theme(legend.position = "bottom") + labs(x = TeX("V + U_0 - U_1"), y = TeX("Y_1 - Y_0"))
```


```{r}
config_4 %>%
  ggplot(aes(x = V_tilde))+ 
  geom_smooth(aes(y = beta_i, color = "MTE (from MC simulation)"), method = "lm") + 
  geom_point(aes(y = beta_i), alpha = 0.1, size = 0.1) +
  geom_line(aes(y = MTE, color = "MTE (theory)")) + labs(color = "") +
  theme(legend.position = "bottom") + labs(x = TeX("V + U_0 - U_1"), y = TeX("Y_1 - Y_0"))
```


## Compute both objective and subjective treatment effects

???

I am assuming the objective effect is just

$$Y_1 - Y_0 $$

## D: Derive weights and estimate treatment parameters

### values estimated from monte-carlo simulations
```{r}

treat_effects <- function(config){
  
  config_num <- filter(config, row_number() ==1)$config_num
  
  
  Prob_treated <- config$D %>% mean()
  
  ATE <- config$beta_i %>% mean()
  
  TT <- filter(config, D == 1)$beta_i %>% mean()
  
  TUT  <- filter(config, D == 0)$beta_i %>% mean()
  
  ols_model <- lm(Y ~ D, data = config)
  
  OLS <- as.numeric(ols_model$coefficients['D'])
  
  #linear IV
  first_stage <- lm(D ~ Z, data = config)
  
  D_hat <- predict(first_stage)
  
  second_stage <- lm(config$Y ~ D_hat)
  
  linear_IV <- as.numeric(second_stage$coefficients['D_hat'])
  
  
  #PRTE?
  
  tibble(config_num,
         Prob_treated,
         ATE,
         TT,
         TUT,
         OLS,
         linear_IV)
  
  
}



map_df(config_list, treat_effects) %>%
  kable(digits = 3, 
        col.names = c("Configuration", "$Pr(D =1)$", 
                      "ATE", "TOT", "TUT", "OLS", "Linear IV"))
```


Taking our expression for

$$\Delta^{MTE} = \bar{\beta} + \frac{-\sigma_1^2 - \sigma_0^2 + 2\sigma_{01} + \sigma_{V1} - \sigma_{V0}}{ \sigma_v^2 + \sigma_0^2 + \sigma_1^2 - 2\sigma_{01} + 2\sigma_{0V} -2\sigma_{1V}}*(V + U_0 -U_1)$$

And subsituting in for $\tilde{V} = \sigma_{\tilde{V}}*\Phi^{-1}(U_D)$, we have

$$\Delta^{MTE} = \bar{\beta} + \frac{-\sigma_1^2 - \sigma_0^2 + 2\sigma_{01} + \sigma_{V1} - \sigma_{V0}}{\sqrt{\sigma_v^2 + \sigma_0^2 + \sigma_1^2 - 2\sigma_{01} + 2\sigma_{0V} -2\sigma_{1V}}}*\Phi^{-1}(U_D)$$

We will label the ratio of the variances

$$c = \frac{-\sigma_1^2 - \sigma_0^2 + 2\sigma_{01} + \sigma_{V1} - \sigma_{V0}}{\sqrt{\sigma_v^2 + \sigma_0^2 + \sigma_1^2 - 2\sigma_{01} + 2\sigma_{0V} -2\sigma_{1V}}}$$

### ATE


$$ATE = \int_{0}^{1} (1)*\Delta^{MTE} du_D$$

So plugging in our expression above we have
$$ATE = \bar{\beta} + c*\int_{0}^{1} \Phi^{-1}(u_d) du_D$$

And note with a change of variables

$$\int_{0}^{1} \Phi^{-1}(u_d) du_D = \int_{-\infty}^{\infty} \tilde{v}*f(\tilde{v})d\tilde{v}$$

Where $f(\tilde{v})$ is the pdf of $\tilde{V} \sim N(0, \sigma_{\tilde{v}})$ as derived in part A. The RHS is just $E[\tilde{V}]= 0$, so we have

$$ATE = \bar{\beta} = 0.2$$
So the ATE should be the same for all the configurations and constant across $u_d$


## TT

$$\Delta^{TT} = \int_{0}^{1} \omega_{TT}*\Delta^{MTE} du_D$$

Where:

$$\omega_{TT} = \frac{Pr(P(z) \geq u_D )}{Pr(D = 1)}$$

Evaluating the denominator
$$E(P(Z)) = E[\Phi(\frac{\bar{\beta}  - Z}{\sigma_{\tilde{V}}})]  $$

Taking the expection w.r.t Z, since $Z ~ U(-0.5, 0.5)$

$$E(P(Z)) = \Phi(\frac{\bar{\beta}}{\sigma_{\tilde{V}}})$$
Note that because $P(z) \sim U(0,1)$, we know 

$$P(P(z) \geq u_D) = 1-u_D$$

We know the 

$$\omega_{TT} = \frac{1 - u_D}{\Phi(\frac{\bar{\beta}}{\sigma_{\tilde{V}}})}$$

So pluggin in our equation for MTE
$$\Delta^{TT}(u_D) = \bar{\beta}*\int_{0}^{1} \omega_{TT} du_D + c*\int_{0}^{1} \Phi^{-1}(u_d)*\omega_{TT} du_D$$



```{r}
weights <- function(config_num,
                     sigma_11, sigma_00, sigma_V, sigma_01, sigma_0V, sigma_1V,
                     beta = 0.2){
  
  u_d <- seq(0, 1, 0.01)
  
  sigma_v_tilde <- sqrt(sigma_00 + sigma_11 + sigma_V -2*sigma_01 +2*sigma_0V - 2*sigma_1V)
  
  tibble(u_d) %>%
    mutate(config = config_num,
           w_ATE = 1,
           w_TT = (1-u_d)/pnorm(beta/sigma_v_tilde))
  
}

all_weights <-
    weights("I", 1,1,1,0,0,0) %>%
  rbind(weights("II", 1,1,1,0.5,-0.5,0),
        weights("III", 1,1,1,0,0,0.5),
        weights("IV", 1,1,0, 0.5, 0, 0),
        weights("V", 1, 0.25, 1, 0.1, -0.2, 0.2),
        weights("VI", 0.25, 0.25, 1, -0.2, 0.1, -0.2),
        weights("VII", 1, 1, 1, -0.2, 0.6, 0.5)) %>%
  pivot_longer(cols = -one_of(c("u_d", "config")), names_to = "parameter", values_to = "weight")


all_weights %>%
  ggplot(aes(x = u_d, y = weight, color = config)) +
  geom_line() + facet_wrap(~parameter)
```


```{r}
treat_effects_theory <- function(config_num,
                     sigma_11, sigma_00, sigma_V, sigma_01, sigma_0V, sigma_1V,
                     beta = 0.2){
  
  sigma_v_tilde <- sqrt(sigma_00 + sigma_11 + sigma_V -2*sigma_01 +2*sigma_0V - 2*sigma_1V)
  
  c <- (-sigma_11 - sigma_00 +2*sigma_01 +sigma_1V - sigma_0V)/sigma_v_tilde^2

  tibble(config_num,
         P_D1 = pnorm(beta/sigma_v_tilde),
         ATE = beta,
        c = c)

}

all_tx_effects <- 
  treat_effects_theory("I", 1,1,1,0,0,0) %>%
  rbind(treat_effects_theory("II", 1,1,1,0.5,-0.5,0),
        treat_effects_theory("III", 1,1,1,0,0,0.5),
        treat_effects_theory("IV", 1,1,0, 0.5, 0, 0),
        treat_effects_theory("V", 1, 0.25, 1, 0.1, -0.2, 0.2),
        treat_effects_theory("VI", 0.25, 0.25, 1, -0.2, 0.1, -0.2),
        treat_effects_theory("VII", 1, 1, 1, -0.2, 0.6, 0.5)) 


all_tx_effects %>%
  kable(digits = 3)
```

```{r,eval= FALSE, include=FALSE}

# working text
and we can evalute the integral

$$\int_0^1 \frac{1 - u_D}{\Phi(\frac{\bar{\beta}}{\sigma_{\tilde{V}}})} du_D = \frac{1}{\Phi(\frac{\bar{\beta}}{\sigma_{\tilde{V}}})}[u_D - (1/2)*u_D^2]_0^1  = \frac{1}{2*\Phi(\frac{\bar{\beta}}{\sigma_{\tilde{V}}})}$$

Plugging this in and using the fact that $\int_{0}^{1} \Phi^{-1}(u_d) du_D = 0$, we have

$$TT = \frac{1}{\Phi(\frac{\bar{\beta}}{\sigma_{\tilde{V}}})}(\frac{\bar{\beta}}{2} - c*\int_{0}^{1} \Phi^{-1}(u_D)* u_D \space du_D)$$


$$TT = \frac{1}{\Phi(\frac{\bar{\beta}}{\sigma_{\tilde{V}}})}(\frac{\bar{\beta}}{2} - c*\int_{-\infty}^{\infty} \frac{\tilde{v}}{\sigma_{\tilde{v}}}* \Phi(\frac{\tilde{v}}{\sigma_{\tilde{v}}}) \space d\tilde{v})$$
```



